---
up---
title: "Monarch Watch"
author: "Schwab"
format: html
editor: source
---

```{r}
#install.packages("googlesheets4")
library(googlesheets4)  #only use google sheets if you are not one the server. 
library(tidyverse)
library(janitor)
library(tidygeocoder)
library(sf)
```

Every year Monarch Butterflies fly south for the winter. They start in higher latitudes and migrate south to their wintering areas in Mexico.

[Monarch Watch](https://monarchwatch.org/blog/) is a citizen science initiative housed in the University of Kansas. Every year volunteers tag these butterflies and release them so their progress in their southern migration can be mapped. If someone finds a monarch with a tag they [can report it.](https://docs.google.com/forms/d/e/1FAIpQLSetTI4HYcyb1GfIyo-U5rAuVVezvOrwicz2ztIt244Vh7S4VA/viewform) We'll examine only the Eastern Monarch's trip south.

## Import and Clean Data

The reporting data is available online in the form of google sheets. It can be brought into R with the `googlesheets4` R package. Pay attention to your console, google needs permission to download.

After importing clean the names of the data frames

```{r}
US_CAN_Recovery_2022 = read_csv("data/us_can_recovery.csv")
MEX_Recover_pre_2023 = read_csv("data/mexico_data.csv")
```

```{r}
#| eval: False
#| message: False

####################### If you are NOT on the server you can run this chunk ############################
####################### If you ARE on the server just run the chunk above   ############################

# US and Canada Data
US_CAN_Recovery_2022 <- read_sheet("https://docs.google.com/spreadsheets/d/14ONbP-0rgvVz-DR0MWYkkPm0BzyFBHMiNCA2F5I8Yks/edit#gid=1710298744") 

# Mexico Recovery Data
MEX_Recover_pre_2023 <- read_sheet("https://docs.google.com/spreadsheets/d/1UdJfooBJrm0Y1zlpwIhGz7ToZfP9h8OeucJbXrWwZEY/edit#gid=1853245517")

# If you cannot get access to the above data run these two lines in the console. Otherwise ignore this message
# googlesheets4::gs4_deauth()
# googlesheets4::gs4_auth()

# clean the names here. 
US_CAN_Recovery_2022 <- US_CAN_Recovery_2022 |> clean_names()
MEX_Recover_pre_2023 <- MEX_Recover_pre_2023 |> clean_names()

write_csv(US_CAN_Recovery_2022, "data/us_can_recovery.csv")
write_csv(MEX_Recover_pre_2023, "data/mexico_data.csv")

```

## North American Map

I found an old map of north america that we can use for this lab.

```{r}
# map data from https://earthworks.stanford.edu/catalog/stanford-ns372xw1938
# I made a North America Map for you. 

north_am = read_sf("data/north_america/")|>
  st_transform(crs = 4326)|> 
  filter(COUNTRY %in% c("MEX", "USA","CAN"))|>
  separate_wider_delim(STATEABB, delim = "-", names = c("country_abb","state_abb") ) |>
  st_as_sf()

```

Plot the map to make sure the states/provinces are shown are shown.

```{r}
# Map out USA_map here. 
ggplot()+
  geom_sf(data = north_am)+
  theme_void()+
  coord_sf(xlim = c(-130, -60), ylim = c(15,60), default_crs = 4326, crs = 3978) # Note the change in projection.
```

## Find top 7 butterflies

I'd like to trace the paths of these butterflies on a map of the US. Each butterfly has a tag. So if the same tag is repeated multiple times we can follow its path.

Wrangle US_CAN_Recovery_2022 to get the top 7 most spotted butterflies 2022. Then use pull() to save the most spotted tag codes as a vector so we can use it to filter later.

```{r}

# wrangle and pull here. 
top_7_most_spotted_2022 <- US_CAN_Recovery_2022  |>
  count(tag_code) |>
  slice_max(n=7, order_by = tag_code) |>
  pull(tag_code)

```

To map the butterfly sightings we need the location data for the top 7 butterflies. `filter()` for the top seven most seen butterflies, pipe that into `select()` to get the location of the city and state and finally pipe that information into `geocode()`. It takes `geocode()` sometime to produce lat and long data. Make sure your filtering works as expected before using `geocode()`. This will give us the lat and long at the city level. To be safe you should use st_as_sf() to give the proper map projection.

```{r}
##### Geocoding is hard on the server. 


top_7_butterfly <- US_CAN_Recovery_2022 |>
# wrangle here. 
  filter(tag_code %in% top_7_most_spotted_2022) |>
  select(tag_code,city_location,state_province) |>
  geocode(city = city_location, state= state_province)|>
  st_as_sf(coords = c("long", "lat"), crs = 4326)
```

In the code chunk below add your top 7 butterflies, be sure to jitter them and make them orange.

```{r}
# Make your top 7 butterfly map here. 
north_am|>
  ggplot()+
  geom_sf() +
  geom_sf(data = top_7_butterfly, size = 2, color="orange")+
  theme_void()+
  coord_sf(xlim = c(-130, -60), ylim = c(15,60), default_crs = 4326, crs = 3978) # Note the change in projection.

```

This is not an exciting map, I only see 7ish dots, not seven butterflies travelling south. I guess one butterfly rarely gets spotted again across state boarders. This shows the limits of citizen science. Not many people participate so the data is spotty.

## Make a Chrolopleth map of the US and Canada.

Let's see what we can learn when we plot all of the butterflies from 2022 at the same time. Let's code it by the state level, as opposed to the city level to save time. Getting location at the state level will take `geocode()` 43 seconds instead of 8 minutes at the city level. Once you geocode your data you can use save() to save the data, and load() to load it back in when you render. You should comment out the geocoding and save function afterward, uncomment the load().

```{r}
#| eval: false

# eval is set to false so this chunk will not execute when you render. 

 butterfly_location_data_state <- US_CAN_Recovery_2022 |>
   select(tag_code,city_location,state_province,country, date)|>
   geocode(state = state_province, country = country)

# This saves the data by state, so you don't have to geocode it everytime. Note that this chunk will not eval when rendered. You should see butterfly_location_data_state.Rdata in the files after running this chunk. 
save(butterfly_location_data_state, file = "butterfly_location_data_state.Rdata")
```

```{r}
# This chunk will load in the saved geolocation of the butterflies. 
load("data/butterfly_location_data_city.Rdata")

```

We would expect locations in the US with more people to have more sightings. So instead of plotting the overall sightings we'll make a Chloropleth map that colors the sightings by state as a percentage of the total sightings. Be mindful and ask yourself: " are we really just creating a population map here? "

We have to wrangle the data a bit. Make a df called `butterfly_location_summary` wrangle to find the following:

-   The total number of sightings per state
-   The total number of sightings in 2022
-   The proportion for each state

Then we need to join our data frame with `USA_map` to get the location data. I made a `states` df that has the states names and abbreviations in it to help with joining.

-   join the states df to the butter_location_summary

-   join the new df made above to USA_map, be sure to have US map on the left so you don't lose simple features.

```{r}
# Wrangle your data
butterfly_location_summary <- butterfly_location_data |>
   group_by(state_province) |>
   summarise(
          count_of_sightings = n(),
          total = nrow(x=butterfly_location_data),
          prop = count_of_sightings/total)


# Here is a df with state abreviation and names
#states<- as.data.frame(cbind(abrev = state.abb, lower_state = tolower(state.name), upper_state = state.name))

#butterfly_location_data_summary <- inner_join(states, butterfly_location_summary, by = c("abrev" = "state_province"))
north_am = read_sf("data/north_america/")|>
  st_transform(crs = 4326)|> 
  filter(COUNTRY %in% c("MEX", "USA","CAN"))|>
  separate_wider_delim(STATEABB, delim = "-", names = c("country_abb","state_abb") )

# Inner join to get the location data.

north_am <- north_am |>
  inner_join(butterfly_location_summary, by = c("state_abb"="state_province")) |>
  st_as_sf()

```

Great! Let's make a map of the number of sightings of butterflies in each state. Also pick the "YlOrRd" color palette, the default one doesn't conjure up images of Monarchs. Make sure darker means more. I added some themes to make things nicer.

```{r}

# Make map here.
north_am|>
  ggplot()+
  geom_sf(aes(fill=prop)) +
  # The code below set a color palette and changes the background color. Orange colors for butterflys and light blue for the background the make the points pop. I also turned off the grid lines. 
  scale_fill_distiller(palette = "YlOrRd", direction = 1) +
  # I added the following themes you can change them or leave them the same.
  theme_void() +
  theme(
    panel.background = element_rect(fill = "azure2",
                                colour = "black",
                                size = 0.5)
    )
  

```

The above map doesn't normalize for the population of each state. Notice that Texas, NY, NJ, Virginia, and Ohio are darker because more people live there. More people that live there means more people that will report butterflies. We basically just made a population map. :( This isn't very helpful. To illustrate this, run the chunk below of the total number of observations to see that the map is the same.

```{r}
north_am|>
  ggplot()+
  geom_sf(aes(fill=count_of_sightings)) +
  
  # The code below set a color palette and changes the background color. Orange colors for butterflys and light blue for the background the make the points pop. I also turned off the grid lines. 
  
  scale_fill_distiller(palette = "YlOrRd", direction = 1) +
  theme_void() +
  theme(
    panel.background = element_rect(fill = "azure2",
                                colour = "black",
                                size = 0.5)
    )
  
```

I'd like a map that shows the proportion of sightings based on the population of each state. Wrangle the data to show the proportion of sightings from each state.

```{r}
### Make population  data here. ###

  
  # The code below set a color palette and changes the background color. Orange colors for butterflies and light blue for the background the make the points pop. I also turned off the grid lines. 
  
  
  theme_void() +
  theme(
    panel.background = element_rect(fill = "azure2",
                                colour = "black",
                                size = 0.5),
    # remove the lat and long with the appropriate arguement.
    )
  
```

**Question: What does the above map show?**

## An Interactive Map

We'll use leaflet to make a map showing the location of all of the sightings in 2022.

These butterflys end up in Mexico for the winter. Using MEX_Recover_pre_2023 data

-   filter for 2022 data

-   add a column that is called Mexico and is the word "Mexico" repeated.

-   add a column that is the count of the number of sightings in each location.

-   Then geocode by state and country.

```{r}
Mexico_locations <- butterfly_dataframes$MEX_Recover_pre_2023 |>
  filter(#)|>
  group_by()|>
  mutate() |>
    # We need to ungroup our data before we geocode it. 
  ungroup()|>
  geocode(state = , country = )



```

I ran the location by city, and it took 8 minutes. You should be able to load in the data, its called "eight_minute_data.Rdata". I've left my code for the geocoding, just so you can see it. You don't need it. Save yourself 8 minutes and don't run the chunk below only to load in the city level sighting data.

```{r}
# butterfly_location_data_8min <- US_CAN_Recovery_2022 |>
#    select(tag_code,city_location,state_province,country, date)|>
#    geocode(city = city_location, state = state_province, country = country)

#save(butterfly_location_data_8min, file = "butterfly_location_data_8min.Rdata")

# Make sure butterfly_location_data_8min shows in the environment after loading.  
load("eight_minute_data.Rdata")
```

Make a map of mexico.

```{r}
Mexico_map <-  maps::map(region = "Mexico", plot = FALSE, fill = TRUE) |> st_as_sf(coords = c("x", "y"), crs = 4326)

```

This bit of code will bring in the image of a butterfly, so we can make our map pretty. You can see this file in your files.

```{r}
butterfly_icon <- makeIcon(
  iconUrl = "monarch-butterfly.png",
  iconWidth = 9.6, iconHeight = 9.6,
  iconAnchorX = 0, iconAnchorY = 0)
```

Add two sets for markers one for the individual sightings in `butterfly_location_data` and one for the Mexico locations, use the big icons for the wintering homes. Figure out which arguments need to go into the addMarkers() functions.

```{r leaflet}
leaflet() |>
  addTiles() |>
  addMarkers(data= , 
             lat = ~#what variable goes here, leave the ~,
             lng = ~#what variable goes here, leave the ~,
             icon = 
             )|>
  
  # Add a popup with the number of butterflies counted for mexico. 
  addCircleMarkers(data = Mexico_locations,
                   popup = ~as.character(#what goes here?))
    
  
```

This map gives an interesting picture of where this type of citizen science is actually happening.