---
title: "CSI-MTH 190: Intro to Data Science"
subtitle: "Lab 4 APIs and the NYTimes"
author: "Your name"
date: "today"
format:
  html:
    embed-resources: true
execute:
    eval: true
    message: false
---

In this lab we will use an API to grab top ten best sellers list from the New York Times. [This is a nice intro to APIs](https://www.dataquest.io/blog/r-api-tutorial/) if you'd like to read it.

Learning Goals:

1.  Get some familiarity with APIs.
2.  Use some simple data to practice with ggplot().
3.  Make a final plot using skills we've learned in class so far.

```{r}
library(tidyverse)
library(jsonlite) # This is for converting the api's json files to a data frame. 
```

## Step 1: Make a Developer Key for the New York Times.

Go to [this website](https://developer.nytimes.com/get-started) follow the instructions to create an account and sign in.

After you sign in to your New York Times Developers account find the get started tab and click it (It'll bring you back to the get-started page above). Follow the instructions to register an app for the Books API. (The first step may be confusing: click you username in the upper right corner and select apps). I labeled my app "lab 4". Enable permissions for the book API.

After you've created an app go back to the [my-apps page](https://developer.nytimes.com/my-apps) and click on the app to find your key. Paste your key below. Put " " or ' ' around it, it needs to be a string.

```{r}
your_key = 'put key here'
```

## Step 2: Make a Call to an API

With this API we need to first prepare a GET request. This is a specially formatted url, that will reach out to the server at the New York Times. You can see how these are formatted for the best seller list [here](https://developer.nytimes.com/docs/books-product/1/overview). Most APIs are similar, but each has its own file structure. No spaces are allowed in the url. I've made best_seller_list() function which will do this for you. There is no documentation so read the comments on the arguments. The date and genre default to get the hardcover fiction books from Feb 19, 2025.

```{r}
# Don't edit this function

best_seller_list = function(date_nyt = '2025-02-19', genre = 'hardcover-fiction', key = your_key ){
  # This function will grab the 15 best sellers from a given date and genre.
  # Args: 
  #    date_nyt : string - in the form year-month-day, must be a Wednesday after 2017, use current for most recent list.
  #    genre : string - book genre, hardcover-fiction see https://www.nytimes.com/books/best-sellers/
  #    key : string - Your developer API string 

  # Format the url as per the nyt examples:
  list_from_nyt = fromJSON(
    paste("https://api.nytimes.com/svc/books/v3/lists/",date_nyt,"/",genre,".json?api-key=",your_key, sep = "")
    )

  # Make data to a dataframe
  best_sell_hardcover = as.tibble(list_from_nyt$results$books)

  return(best_sell_hardcover)
  }

```

## Call the function

```{r}
# call the function here and save it to a variable called feb_19_25_hcf_books
# this variable will be your data frame. I do this in the intro video. 
feb_19_25_hcf_books = best_seller_list()

```

# New York Times books reviews

Each Wednesday the NYT puts out its best seller lists, each of which is 15 books long. You can see an example [here](https://www.nytimes.com/books/best-sellers/2025/02/19/hardcover-fiction/). In fact that is the list that you got from the API call above and is now saved in feb_19_25_hcf_books. We can ask the API for other book lists, or authors too, and you will do that later. The data is fairly small and not challenging to understand, the rest of the focus will be on showing the data with different graphing techniques.

## Step 3: View the Data Using Two Approaches

View the data frame you have created.

```{r}
# Write your code below
head(feb_19_25_hcf_books)

```

```{r}
# Write your code below
view(feb_19_25_hcf_books)

```

## Step 4: Create a Histogram Visualizing the Distribution of the weeks_on_list Variable.

(1) Be sure to add labels to your plot.
(2) Set appropriate `binwidth` for bins. Don't use the default.
(3) Set the outline color for bins by adjusting the color argument.

```{r}
# Write your code below 
feb_19_25_hcf_books|>
  ggplot()+
  geom_histogram(
    aes(x = 
        ),
    binwidth = ,
    color = ""
    )

```

**Discussion**: How do you interpret this plot? \> Interpreting histograms: (1) Peaks; (2) Spread; (3) Skewed data; (4) Outliers.

**Answer**:

## Step 5: Facet Your Previous Plot by Publisher or Author.

Set `nrow` of your choice in the `facet_wrap` function.

```{r}
# Paste the code below and adjust it 


```

**Discussion**: What do the distributions of the bars indicate? What do you learn when comparing distributions across books?

**Answer**:

## Step 6: The y-axis is too busy.

The y-axis doesn't tell us much information, its just a count of books, which is always just 1. We don't need to to show 0.25, 0.50, or 0.75; remove them by adding the layer: scale_y_continuous( breaks = c( 0, 1)). The clean up your graph by adding a title, changing the labels and applying a theme.

## Step 7: Make Another API call for a different genre and date.

```{r}
# You can reuse the best_seller_list() function, but change the date and genre, be sure to update the data frame name to one that fits your data.


```

## Step 8: Make a final plot

In this last step make a final plot of your choosing using the data from the call in step 6, it can be a plot you made previously or something different. Clean it up, don't use default colors, add context and give good titles and labels. Cite your source. Finally apply a theme to lose the gray background. It can be any theme available (i.e. minimal, bw, classic, etc)

Then write a sentence about something noticeable about the books from your date and genre.

**Answer:**