---
title: "lab 9: SQL Queries "
author: "Solutions: Schwab"
date: "today"
format:
  html:
    embed-resources: true
    code-fold: true
execute: 
  warning: false

---

## Introduction

You have learned a lot about manipulating and visualizing data in R. You have gotten a little practice doing it in SQL. In this lab you will continue that practice in SQL. Specifically you will: 

1. Become familiar with the Chinook database. 
2. You will answer a five questions using SQL.
3. You will formulate your own question and answer it using the Chinook Data. This will also be done in SQL. 


## Check the readme

Before you being he questions, take some time to get to know the [Chinook data](https://github.com/lerocha/chinook-database/tree/master)

# Quesions

There are five questions you must answer about the Chinook Data.

You should answer all questions with your query. The LIMIT and ORDER BY commands will be useful. You should have no extra output. 

```{r}
library(DBI)
mydb <- dbConnect(RSQLite::SQLite(), "course-materials/_labs/lab_9/chinook.sqlite")

```

```{r}
# list the tables
dbListTables(mydb)
```

```{r}
#list the variables within the Track table and what type they are. 
dbGetQuery(mydb, "PRAGMA table_info('Track')")

```

## Question 1

How many artists have more than one album in this data? Which artist has the most albums? (Hint: My answer had more than 50 artist, but less than 60)

```{r}
dbGetQuery(mydb, 
    "SELECT Name, Count(Name) As Count
    FROM Artist
    JOIN Album ON Artist.ArtistId = Album.ArtistId
    Group By Name
    HAVING Count > 1
    ORDER BY Count 
")


```

## Question 2

From which group is the longest song in this data? How long is that song in seconds? What is the Name of the Track? (Hint: I needed two joins, and the "artist" is not a traditional music artist.)

```{r}
dbGetQuery(mydb, 
    "SELECT Artist.Name AS Name, Max(Milliseconds/1000) As seconds , Album.Title, Track.Name
    FROM Track
    JOIN Album on Album.AlbumId = Track.AlbumId
    JOIN Artist on Artist.ArtistId = Album.ArtistId
    
")
```

```{r}
dbGetQuery(mydb, 
    "SELECT Album.Title, Track.Name, Composer
    FROM Track
    JOIN Album on Album.AlbumId = Track.AlbumId
    JOIN Artist on Artist.ArtistId = Album.ArtistId
    
")

```

## Question 3

What is the name and EmployeeId of employee the helped the most customers? (Hint: You should only have 1 row in your answer, not 3 or 8).

```{r}
dbGetQuery(mydb, 
    "SELECT Employee.FirstName,Count(Employee.FirstName), SupportRepId
    FROM Employee
    JOIN Customer ON EmployeeId = SupportRepId
    Group By EmployeeId
    LIMIT 1
")
```

## Question 4

Which Customers paid the most money? Where are they from? (Hint: I found 8 of them).

```{r}
dbGetQuery(mydb, 
    "SELECT FirstName, LastName, Invoice.CustomerId, SUM(Total)
    FROM Invoice
    JOIN Customer  ON   Invoice.CustomerId =Customer.CustomerId
    GROUP BY Customer.CustomerId 
    ORDER BY SUM(Total) DESC
    
    
")
```

## Question 5

Which tracks were purchased the most? Who was the artist of that track? How many times was it purchased? ().

```{r}
dbGetQuery(mydb, 
    "SELECT  Artist.Name, Track.Name, Title, Count(Title) As count
    FROM Customer
    JOIN Invoice ON Customer.CustomerId = Invoice.InvoiceId
    JOIN InvoiceLine ON Invoice.InvoiceId = InvoiceLine.InvoiceId
    JOIN Track ON Track.TrackId = InvoiceLine.TrackId
    JOIN Album ON Album.AlbumId = Track.AlbumId
    JOIN Artist on Artist.ArtistId = Album.ArtistId
    GROUP BY Track.Name
    ORDER BY count DESC
")
```

# Develop your own question

When developing your own question, you must do at least one join and use some kind of aggragting or filter operation. 

## Your question below