---
title: "lab 9: SQL Queries "
author: "Solutions: Schwab"
date: "today"
format:
  html:
    embed-resources: true
    code-fold: true
execute: 
  warning: false

---

## Introduction

You have learned a lot about manipulating and visualizing data in R. You have gotten a little practice doing it in SQL. In this lab you will continue that practice in SQL. Specifically you will: 

1. Become familiar with the Chinook database. 
2. You will answer a five questions using SQL.
3. You will formulate your own question and answer it using the Chinook Data. This will also be done in SQL. 


## Check the readme

Before you being he questions, take some time to get to know the [Chinook data](https://github.com/lerocha/chinook-database/tree/master)

# Quesions

There are five questions you must answer about the Chinook Data.

You should answer all questions with your query. The LIMIT and ORDER BY commands will be useful. You should have no extra output. 

```{r}
library(DBI)
mydb <- dbConnect(RSQLite::SQLite(), "put the path to the chinook data here")

```

```{r}
# list the tables
dbListTables(mydb)
```

```{r}
#list the variables within the Track table and what type they are. 
dbGetQuery(mydb, "PRAGMA table_info('Track')")

```

## Question 1

How many artists have more than one album in this data? Which artist has the most albums? (Hint: My answer had more than 50 artist, but less than 60)

```{r}
dbGetQuery(mydb, 
    "
    ")


```

## Question 2

From which group is the longest song in this data? How long is that song in seconds? What is the Name of the Track? (Hint: I needed two joins, and the "artist" is not a traditional music artist.)

```{r}
dbGetQuery(mydb, 
    "
    ")
```

## Question 3

What is the name and EmployeeId of employee the helped the most customers? (Hint: You should only have 1 row in your answer, not 3 or 8).

```{r}
dbGetQuery(mydb, 
    "SELECT Employee.FirstName,Count(Employee.FirstName), SupportRepId
    FROM Employee
    JOIN Customer ON EmployeeId = SupportRepId
    Group By EmployeeId
    LIMIT 1
")
```

## Question 4

Which Customers paid the most money? Where are they from? (Hint: I found 8 of them).

```{r}
dbGetQuery(mydb, 
    "
    ")
```

## Question 5

Considering the 8 customers from question 4, which track was purchased the most? Who was the artist of that track? How many times was it purchased? (Hint: Recycle the code from problem 4, I did 5 joins for this problem).

```{r}
dbGetQuery(mydb, 
    "
    ")
```

# Develop your own question

When developing your own question, you must do at least one join and use some kind of aggragting or filter operation. 

## Your question below