---
title: "Taxonomy of Graphs part 2"
subtitle: "CSI-MTH-190"
author: "Schwab"
editor: visual
format: 
  revealjs: 
    theme: beige
execute: 
  echo: true
---

## PreReqs

I assume you've read chapter 2.2, if you haven't, go back and read it before this lecture. Also I assume you saw the previous lecture on taxonomy.

## The Data

Here's the starwars data again. 

```{r}
library(tidyverse)
library(mdsr)
head(starwars)
```

##  {background-image="https://s-media-cache-ak0.pinimg.com/originals/c2/46/92/c24692fd95dbfb89f772613a316a5405.png" background-size="contain"}

## Visual Cues Position

Show position with a point plot. You will need to map two asthetics. 

```{r}
starwars |>
    ggplot()+
    geom_point(aes(x = height, y = mass))
```

## Visual Cues position again

I'm going to remove the really heavy character with a filter. 

```{r}
starwars |>
    filter(mass < 250)|>
    ggplot()+
    geom_point(aes(x = height, y = mass))
```

## Visual Cues Shape

We can map the categorical variable shape. 

```{r}
starwars |>
    filter(mass < 250)|>
    ggplot()+
    geom_point(aes(x=height, 
            y = mass,
            shape = gender
            ))
    
```

## Visual Cues Shape Clean

We can map the categorical variable shape. 

```{r}
starwars |>
    filter(mass < 250)|>
    ggplot()+
    geom_point(aes(x=height, 
            y = mass,
            shape = gender
            ))+
    theme_minimal()
    
```

## You try: 

Take a moment and copy the code below into r. Change the x-axis to be birth_year, and the y-axis to be height. Are older characters taller? Color by sex, what do you notice?

```{r}
#| eval: False

starwars |>
    filter(mass < 250)|>
    ggplot()+
    geom_point(aes(x= , 
            y =  ,
            color = 
            ))+
    theme_minimal()
```


##  {background-image="https://i.pinimg.com/736x/62/c6/ee/62c6ee636c47dccad00e7e56c268c3bf--data-visualization-points.jpg" background-size="contain"}

## Visual Cues Context

Here I add a title, x and y axes and a source of the data. 

```{r}

starwars |>
    filter(mass < 250)|>
    ggplot()+
    geom_point(aes(x = birth_year, 
            y = height,
            color = sex
            ))+
    labs( x = "year character was born",
        y = "height (cm)" ,
        title = "Starwars Characters",
        caption = "source: https://swapi.py4e.com/"
        )+
    theme_minimal()
```

