---
title: "Joins"
subtitle: "CSI-MTH-190"
author: "Schwab"
format: 
  revealjs:
    theme: beige
execute: 
  warning: false
  message: false
  echo: true
--- 

```{r}
library(tidyverse)
```

## Combine two data frames 

- bind_cols()
- inner_join()
- left_join()
- full_join()

## Our toy data frames

```{r}
band_members
#band_members2
band_instruments
```

## bind_cols() 

This function just smashes two dataframes together. 

Only use if you are sure that the rows match up together. 

```{r}
bind_cols( band_instruments, band_members)
```

## full_join()

Keeps all the data and by default matches by a column name. 

```{r}
full_join(band_instruments, band_members)
```

## left_join()

Keeps all data in the left dataframe, adding rows from the right dataframe if there is a key match. 

```{r}
left_join(band_instruments, band_members)
```

## inner_join()

Keeps only the data when keys match in both data frames. This is helpful when you want to know what values the dataframes have in common. 

```{r}
inner_join(band_instruments, band_members)
```

## Different Key names

If the keys in both data frames after different we must specify how to join the data. 

```{r}
left_join(
    band_instruments2, 
    band_members, 
    by = c("artist" = "name")
    )
```

## other joins

There are many other joins, some of which are on [dplyr's cheatsheet](https://github.com/rstudio/cheatsheets/blob/main/data-transformation.pdf)

