---
title: "Welcome to Data Science "
subtitle: "CSI-MTH-190: Topics"
author: "Schwab"
editor: visual
format: 
  revealjs: 
    theme: beige
    code-fold: true
execute: 
  echo: true

---

## Introduction to Programming Language for Data Science (R) and Computing Environment (RStudio)

-   Take a look at R Studio together

-   [Short History of R](https://nics-github.github.io/SDS192/course-materials/lectures/2_Brief-R-history.html)

## Data Science Life Cycle

[DS life cycle](https://towardsdatascience.com/data-science-101-life-cycle-of-a-data-science-project-86cbc4a2f7f0)

## Grammar of Graphics

[Hadley Wickham](https://byrneslab.net/classes/biol607/readings/wickham_layered-grammar.pdf)

## Web Scraping and Importing Data 

The follow data was taken from [wikipedia](https://en.wikipedia.org/wiki/List_of_best-selling_game_consoles) on May 12, 2026.

```{r}
library(tidyverse)
library(rvest)

video_game_sys = read_html("https://en.wikipedia.org/wiki/List_of_best-selling_game_consoles") |>
html_element("#mw-content-text > div.mw-content-ltr.mw-parser-output > table")|>
html_table() 

video_game_sys |> head()
```

## ( a bit of cleaning)

Data is Dirty we have to start by cleaning it up. 

```{r}
library(janitor)
video_game_sys = video_game_sys |>
  clean_names()|>
  mutate(units_sold= parse_number(units_sold))|>
  select(-ref)

video_game_sys |> head()
```

## Data Wrangling

Here we filter for just the big three: Nintendo, Sony, Microsoft.

```{r}

video_game_sys = video_game_sys |>
  filter( company %in% c("Sony","Nintendo","Microsoft"))

video_game_sys |> head()
```

## Data Visualization

This is a bar chart of the number of video game consoles sold by each company. 

```{r}
# make simple plot of console data. 

video_game_sys |>
  ggplot()+
  geom_col(aes(x = company, y = units_sold), fill = "orange")+
  labs( title = "Number of Systems Sold by Company",
  y = "units sold (millions)",
  caption = "wikipedia: List_of_best-selling_game_consoles")+
  theme_minimal()

```


## Data Management

```         
-   Keeping yourself organized!
-   How Data is housed
```


Data Source: wikipedia

Date: May 12, 2026

We'll also find relevant data online. 

## Data Ethics

How we collect and use data responsibly.

Consider Personally Identifyable Information. (PII)

Who is collecting the data, and for what purpose?

Is the data being misconstrued (intentionally or not).

## Introduction to Database Querying with SQL

Wrangle some data with the classic SQL.

[Simple Query?](https://www.sql-practice.com/)

## Geospatial Data and Maps

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false

library(tidyverse)
library(tidycensus)

#tidycensus::census_api_key("4acec63df8eee0def6e28f09f8ae46d8cc229184", install = TRUE)
options(tigris_use_cache = TRUE)
# Write your code below
mass_prop_retirees <- 
  tidycensus::get_acs(
    geography = "county", 
    variables = c(population = "B01003_001",age_1="B01001_044", age_2='B01001_045', age_3='B01001_046',age_4='B01001_047', age_5 = 'B01001_048', age_6 = 'B01001_044'),
    state = "MA",
    geometry = TRUE,
    cache_table = TRUE
  )|>
  # Add centroids to each region using purrr package
 tidyr::separate(NAME, c("County", "State"), sep = ", ") |> 
 tidyr::separate(County, c("County", "Fluff"), sep = " ") 

mass_prop_retirees <- mass_prop_retirees |> 
  select(County,variable,estimate)|>
  pivot_wider(names_from = variable, values_from = estimate) |>
  mutate(retire_prop = (age_1+age_2+age_3+age_4+age_5)/population)

ggplot() +
  geom_sf(data = mass_prop_retirees, aes(fill = retire_prop)) +
  scale_fill_distiller(direction = 1)+
  labs(title = "Proportion of People over 65",
  caption = "American Community Survey, 2023",
  fill = "prop")+
    theme_void() +
    scale_fill_distiller(palette = "Greys", direction = 1)

```



