---
title: "Web Scrapping"
subtitle: "and cleaning"
author: "Schwab"
format: 
  revealjs:
    theme: beige
editor: visual
execute: 
  echo: true
  code-fold: true
---

## Ways to get data

We have looked at several places to find data:

packages

csv files

APIs

In this lecture we will consider scraping.

## How does it work?

Every web page is an html file.

html stands for HyperText Markup Language.

html has markers to tell the web browser how to display the page.

There are several markers that could be used for scraping.

We will focus primarily on the \<table\> marker

## How to see markers

You can right click on a website an select "inspect" to view a page's html file.

We'll consider the Wikipedia page for the 2026 Winter Olympics.

There are a lot of tables on this site.

<https://en.wikipedia.org/wiki/2026_Winter_Olympics>

## rvest

We will scrape with a package called Rvest.

rvest (meant to sound like "Harvest") is a package that is part of the tidyverse, but is not loaded by default.

This means you download when you download the tidyverse.

But you must still load it after you have loaded the tidyverse.

```{r}
library(tidyverse)
library(rvest)
```

## The read in the html

There are two helpful examples [here](https://rvest.tidyverse.org/).

Generally we will first read in the html data with `read_html()`

`read_html()` takes a url as a string and you have to be connected to the internet for it to work properly. 

```{r}
winter_olympics = read_html("https://en.wikipedia.org/wiki/2026_Winter_Olympics")

winter_olympics
```

## Find the right table

We want to find the "selector" for our table.

```{r}
winter_olympics |>
html_element("#mw-content-text > div.mw-content-ltr.mw-parser-output > table:nth-child(85)")
```

## Convert into a table

We use html_table() to create a data frame. 

```{r}
winter_olympics |>
html_element("#mw-content-text > div.mw-content-ltr.mw-parser-output > table:nth-child(85)")|>
html_table()
```

## Cleaning

Note: We'd have to clean this table a fair bit. Its not tidy and there are strange symbols.

Also everything is a character vector.

We will not clean this table but we'll clean the next one.

## Medal count

Attempt to scrape the 2026 Winter Olympics medal table. We'll clean this one up a bit.


## Medal Count Solution

```{r}

winter_olympics = read_html("https://en.wikipedia.org/wiki/2026_Winter_Olympics")

olympic_medals = winter_olympics |>
  html_element("#mw-content-text > div.mw-content-ltr.mw-parser-output > table.wikitable.sortable.sticky-header-multi.plainrowheaders.jquery-tablesorter")|>
  html_table()
```

## Cleaning with Janitor

We can clean this table with the `dplyr` and `janitor` packages. `janitor` is not part of the `tidyverse`, but it works with tidy data.

See [this page for more information.](https://cloud.r-project.org/web/packages/janitor/vignettes/janitor.html)

## Items to clean

Try some of these yourself first, see the hint functions for each line: 

- Rank should be an integer, not a character  (mutate())

- .mw-paser... should say "name_of_country"  (rename())

- all column names should be lowercase to match style (clean_names())

- lose the last "total row" (head())

- should probably change "remaining" to "other" (case_when(), or manually)

- drop the "total" column (select())

- table is not tidy (pivot_longer())

## Rank should be an integer, not a character

Use mutate to make a new column of the correct type.

```{r}
olympic_medals = olympic_medals |>
  mutate(Rank = as.integer(Rank))
  olympic_medals
```

## .mw-paser... should say "name_of_country"

We'll use rename()

```{r}
olympic_medals = olympic_medals|>
  rename( name_of_country = `.mw-parser-output .tooltip-dotted{border-bottom:1px dotted;cursor:help}NOC`)
```

## all column names should be lowercase to match style (clean_names())

The clean_names() function will lower case all column names and use underscores as spaces. This conforms to the tidyverse style guide. It also makes data easier to work with. 

```{r}
library(janitor)

olympic_medals = olympic_medals |> 
  clean_names()
```

## lose the last "total row" (head())

```{r}
olympic_medals = olympic_medals |>
head(16)

```

## change "remaining" to "other" (case_when(), or manually)

```{r}
olympic_medals|>
mutate(name_of_country = case_when(
  name_of_country=="Remaining" ~ "Other Countries",
  TRUE ~ name_of_country))

# alternative method
#olympic_medals[16, "name_of_country"]<-"hello" # Notice the <- here. 16th row of the name_of_country column.

```

## drop the "total" column

```{r}
olympic_medals = 
  olympic_medals |>
  select(-total)
```

## tidy up the table

```{r}
olympic_medals = olympic_medals |>
pivot_longer(
  cols = c(gold,silver,bronze),
  names_to = "medal", 
  values_to = "number_of_medals")
```

## admire the clean data frame

```{r}
head(olympic_medals)
```