---
title: "Chloropleth and Population Maps"
author: "Schwab"
format: 
  revealjs:
    theme: beige
editor: visual
execute: 
  echo: true
  warning: false
  code-fold: true
---

```{r, echo=FALSE}
# Load all packages here. It's good practice to load all packages at the
# beginning of a qmd file so that collaborators can see in one shot all packages
# they'll need installed:
library(tidyverse)
library(rvest) # This is for scraping VT ski areas
library(janitor) # cleaning

# Libraries for Mapping
library(sf) # shape files
library(maps) # we get the maps of VT and MA from here. 
library(tidygeocoder) # geocoding. 
library(tidycensus)
census_api_key("4acec63df8eee0def6e28f09f8ae46d8cc229184")
```

## Do not accidentally make population maps

Let's make a population map of Mass by county

We need to weight our data in some way.

```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false

# Where do all the under 5 year olds live?

mass_pop <- 
  tidycensus::get_acs(
    geography = "county", 
    variables = c(population = "B01003_001", under_5m="B01001_003", under_5f= "B01001_027"),
    state = "MA",
    # We'll need to pivot wider, but that doesn't work with simple features. 
    geometry = TRUE
  )|>
  # Add centroids to each region using purrr package
 tidyr::separate(NAME, c("County", "State"), sep = ", ") |> 
 tidyr::separate(County, c("County", "Fluff"), sep = " ") 

under_5 = mass_pop |> 
  select(County,variable,estimate)|>
  pivot_wider(names_from = variable, values_from = estimate)|>
  group_by(County)|>
  reframe(County, pop = population, tot_udr_5 = under_5f+under_5m, prop_udr_5 = (under_5m+under_5f)/population)

mass_pop_under_5 = left_join(mass_pop, under_5, by= "County") |>
  select(-c(variable,estimate,moe))|>
  distinct()

```

```{r}
#| eval: false
#| echo: false
st_write(mass_pop_under_5, "course-materials/lectures/mass_pop_under_5.shp")

```

```{r}
#| code-fold: true
#| label: A population map


mass_pop_under_5 = read_sf("mass_under_5")

ggplot() +
  geom_sf(data = mass_pop_under_5, aes(fill = pop)) +
  scale_fill_distiller(direction = 1)+
  labs(title = "Population under 5",
  caption = "American Community Survey, 2023")+
    theme_void()

```

## Graph the Under 5s

```{r}
#| code-fold: true
#| label: Another population map

ggplot() +
  geom_sf(data = mass_pop_under_5, aes(fill = tot_udr_5)) +
  scale_fill_distiller(direction = 1)+
  labs(title = "Children under 5",
  caption = "American Community Survey, 2023")+
    theme_void()
```

## Normalize the Data

One way to do this is:

Divide the totals under 5 by the total population per county for proportion per county.

```{r}
#| code-fold: true
#| label: Not a population map

ggplot() +
  geom_sf(data = mass_pop_under_5, aes(fill = prop_udr_5)) +
  scale_fill_distiller(direction = 1)+
  labs(title = "Proportion of Children under 5",
  caption = "American Community Survey, 2023")+
    theme_void()
```

## Exercise

Load in the mass_retirees data. Make two maps:

-   One which graphs retiree population per county,
-   one which graphs the proportion of retirees per county.

Which counties in Massachusetts have the highest proportion of retirees ?

```{r}
#| echo: false
#| eval: false
#| include: false
#| warning: false

library(tidyverse)
library(tidycensus)

# Write your code below
mass_prop_retirees <- 
  tidycensus::get_acs(
    geography = "county", 
    variables = c(population = "B01003_001",age_1="B01001_044", age_2='B01001_045', age_3='B01001_046',age_4='B01001_047', age_5 = 'B01001_048', age_6 = 'B01001_044'),
    state = "MA",
    geometry = FALSE
  ) |>
  # Add centroids to each region using purrr package
 tidyr::separate(NAME, c("County", "State"), sep = ", ") |> 
 tidyr::separate(County, c("County", "Fluff"), sep = " ") 

mass_prop_retirees <- mass_prop_retirees |> 
  select(County,variable,estimate)|>
  pivot_wider(names_from = variable, values_from = estimate) |>
  mutate(retire_prop = (age_1+age_2+age_3+age_4+age_5)/population)

# Do a left join, a right join will lose the multipolygon-ness
mass_prop = left_join(mass_pop, mass_prop_retirees, by= "County") |>
  select(-c(Fluff,variable,estimate,moe, age_1:age_5), retiree = population)|>
  distinct()
```

```{r}
#| eval: false
#| echo: false

st_write(mass_prop, "course-materials/lectures/mass_retiree/mass_retiree.shp")
```

```{r}
#| code-fold: true

mass_retirees = read_sf("mass_retiree")

```

## 

## Solutions

```{r}
#| echo: false

ggplot() +
  geom_sf(data = mass_retirees, aes(fill = retiree)) +
  scale_fill_distiller(direction = 1)+
  labs(title = "Number of People over 65",
  caption = "American Community Survey, 2023")+
    theme_void()

ggplot() +
  geom_sf(data = mass_retirees, aes(fill = rtr_prp)) +
  scale_fill_distiller(direction = 1)+
  labs(title = "Proportion of People over 65",
  caption = "American Community Survey, 2023")+
    theme_void()  
```