---
title: "Basic Statistics"
subtitle: "CSI-MTH-190"
author: "Schwab"
editor: visual
format: 
  revealjs: 
    theme: beige
execute: 
  echo: true
---

# Statistics

If you've taken statistics before this will be a bit of review:

I'll briefly discuss these topics and how to compute them in R.

-   Measures of Center

-   Sample Size

-   Population Size

-   Measures of Spread

## Measures of center

Often we want to know the average of some data.

There are three common ways of doing this.

mean - sum of all values divided by the number of values.

median - the number in the middle of the values.

mode - the most common number.

## The data

Let's make some data. Pretend I went out and asked 4 people who live in Holyoke, MA their salary. Below is a summary of the answers:

| person | salary  |
|--------|---------|
| 1      | 40000   |
| 2      | 50000   |
| 3      | 60000   |
| 4      | 1000000 |

: fictional annual salaries from 4 people

To bring these salaries into R we can create a vector with the c() function, and save it to a variable called salaries

```{r}
salaries = c(40000,50000,60000,1000000) 
```

##  

## Mean of the Data

We can find the mean of the data with the `mean()` function.

```{r}
mean(salaries)
```

In this case r adds up the values and divides by 4.

The mean of the data, in this case, is not a good representation of the people surveyed.

The mean is sensitive to outliers.

## Median of the Data

To find the median we use the `median()` function.

```{r}
median(salaries)
```

In this case R finds the number in the middle of the 4. Note there is no number exactly in the middle so R finds the mean of the two closest to the middle (50,000 and 60,000).

In this case the median is a better representation of most of the people we surveyed.

The Median is not sensitive to outliers.

## The Mode

There is no mode function in R, and usually we find the mode just by inspection of the data. We can use the `table()` function for this.

```{r}
table(salaries)
```

We can see this data has no mode.

## Sample size

The sample size of our data is the number of observations made. An observation in this case is a person's salary. For our data above the sample size is 4.

## Population size

The population is the total size of the thing you are studying. So the population of the salary data may be everyone that lives in Holyoke, MA.

## Measures of Spread

There are two measures of spread that I'll mention from time to time in this course.

Inner Quartile Range. This is the width between 25% and 75% of the data. We can find it with the `IQR()` function

Standard Deviation. This is the average distance from the mean of the data. We calculate it with the `sd()` function.

We'll discuss these two more as they come up in the course.