---
title: "Taxonomy of Graphs part 1"
subtitle: "CSI-MTH-190"
author: "Schwab"
editor: visual
format: 
  revealjs: 
    theme: beige
execute: 
  echo: true
---

## PreReqs

I assume you've read chapter 2.2, if you haven't, go back and read it before this lecture. 

## The Data

In this lecture I'll use one set of data called `starwars` to illistrate most of the ideas you should consider when building a graph. 

We'll focus mostly on the categorical variables in this lecture. Next week we'll tackle the numeic variables. 

```{r}
library(tidyverse)
library(mdsr)
head(starwars)
```

##  {background-image="https://s-media-cache-ak0.pinimg.com/originals/c2/46/92/c24692fd95dbfb89f772613a316a5405.png" background-size="contain"}

## Visual Cues Length

Show length with a bar chart

```{r}
starwars |>
    ggplot()+
    geom_bar(aes(x=hair_color))
```

## Visual Cues Length again

Show length with a bar chart

```{r}
starwars |>
    ggplot()+
    geom_bar(aes(x=hair_color))+
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

## Visual Cues Color

Too many colors is not helpful. Below I fill each rectangle by `hair_color`.

And we now have a redundant mapping.

```{r}
starwars |>
    ggplot()+
    geom_bar(aes(x=hair_color, fill = hair_color))+
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

## Visual Cues Color again

Limit your colors to show another variable. Here I'll fill by sex.  

```{r}
starwars |>
    ggplot()+
    geom_bar(aes(x=hair_color, fill = sex))+
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

## You try: 

Take a moment and copy the code below into r, fill by eye_color. What do you think about the graph? 

```{r}
#| eval: False

starwars |>
    ggplot()+
    geom_bar(aes(x=hair_color ))+
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

##  {background-image="https://i.pinimg.com/736x/86/a4/93/86a49342761ca36aeca5f8ee03c64867--data-visualization-points.jpg" background-size="contain"}

## Visual Cues Scale

In the graph below the scale is  categorical on the x-axis and numeric on the y-axis. 

```{r}
starwars |>
    ggplot()+
    geom_bar(aes(x=hair_color, fill = sex))+
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

##  {background-image="https://i.pinimg.com/736x/62/c6/ee/62c6ee636c47dccad00e7e56c268c3bf--data-visualization-points.jpg" background-size="contain"}

## Visual Cues Context

Here I add a title, x and y axes and a source of the data. 

```{r}
starwars |>
    ggplot()+
    geom_bar(aes(x=hair_color, fill = sex))+
    labs(x = "Hair Color",
        y = "number of characters",
        title = "Hair Color of Starwars Characters",
        caption = "data from dplyr r package")+
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

## Touch ups

You can add a theme to the graph to easily make things look more professional. Pick your favorit theme and go with it. 

Also I eliminated the x-axis label as it was redundant with the title. 

```{r}
starwars |>
    ggplot()+
    geom_bar(aes(x=hair_color, fill = sex))+
    labs(x = "Hair Color",
        y = "number of characters",
        title = "Hair Color of Starwars Characters",
        caption = "data from dplyr r package")+
    theme_mdsr()+
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

