---
title: "Non-lab: Static Maps w/ ggplot"
author: "Schwab"
format: 
  revealjs: 
    theme: beige
editor: visual
execute: 
  echo: true
  warning: false
  message: false
---

This is not a lab, it is a series of examples to walk you through static mapping in R. Watching the maps video for the week will be helpful here. I recommend you create a RProject file for this document. This will help you manage your working directory.

```{r, echo=FALSE}
library(tidyverse)
library(fivethirtyeight) # This is the presidential data
library(rvest) # This is for scraping VT ski areas
library(janitor) # cleaning
library(leaflet)
library(macleish)

# Libraries for Mapping
library(sf) # shape files
library(maps) # we get the maps of VT and MA from here. 
library(tidygeocoder) # geocoding. 

```

# Tip {.unnumbered}

Do not run entire code blocks at once, as you'll skip over a lot of very important details. Rather, I recommend you highlight small segments of code and run them individually using:

-   macOS: Command + Enter
-   Windows: Control + Enter

# 1. Converting data frames to sf objects

# Example together

```{r}
# Convert pres_2016_trail data frame to sf object using st_as_sf() from sf
# package. Note we set the coordinate reference system to WGS84
# https://spatialreference.org/ref/epsg/wgs-84/
pres_2016_trail_sf <- pres_2016_trail |>
  st_as_sf(coords = c("lng", "lat"), crs = 4326)
pres_2016_trail_sf

# Load maps package map object of US states and plot using default R plot
# function (not ggplot2). Plot is not pretty!
USA_map <-  maps::map("state", plot = FALSE, fill = TRUE)
plot(USA_map)

# Convert maps package map object to sf object using st_as_sf():
USA_sf <- USA_map |> 
  st_as_sf(coords = c("x", "y"), crs = 4326)

# Plot both layers:
ggplot() +
  geom_sf(data = USA_sf) +
  geom_sf(data = pres_2016_trail_sf, aes(color = candidate), size = 2) +
  facet_wrap(~candidate)
```

## Example together

Together: Draw a map of the US with the 3 colleges I've worked at marked:

```{r}
# How did I get the longitude and latitude for Worcester Polytechnic Institute and Smith College? By tapping into the OpenStreetMap (OSM) database using the geocode_OSM() function from the
# tmaptools package. Be careful though, apparently you are limited to 2000
# queries a day.

colleges <- tibble(
  name = c("Holyoke Community College", "Smith College", "Appalachian State University"), 
  long = c(-72.6527, -72.640, -81.8643),
  lat = c(42.1953, 42.316, 36.2136)
)

# Plot using ggplot
ggplot()+
  geom_sf(data = USA_sf)+
  geom_point(data = colleges, aes(long,lat))


# Plot using geom_sf and simple features
colleges = colleges |>
  st_as_sf(coords = c("long", "lat"), crs = 4326)

# geom_sf figures out the simple features and plots them
ggplot()+
  geom_sf(data = USA_sf)+
  geom_sf(data = colleges)


```

## Exercise

Draw a map of only Massachusetts with Boston, and Holyoke marked. This will require you to manually create a data frame like `colleges` above. Alternatively you can use the geocode() function from the lecture on maps, or just google each of the places and make a data frame.

```{r}
# Nothing to do here, I made a map of Mass for you using the maps library

ma_counties <- maps::map("county",regions = "massachusetts",  plot = FALSE, fill = TRUE) |>
  # This sets the simple features 
  st_as_sf(coords = c("x", "y"), crs = 4326) |>
  # This adds a new column with the names of the counties. 
  mutate(county_name = str_replace(ID,pattern = "massachusetts,", replace = ""))
```

```{r}
# Write your code below

  
```

# 2. Together: Loading shapefiles into R

Let's load a shapefile of all bike trails in Massachusetts as provided by [MassGIS (Bureau of Geographic Information)](https://www.mass.gov/info-details/massgis-data-bicycle-trails)

You'll need to upload the biketrails_arc file into your R project folder and edit the path as necessary.

```{r}
# Load the "shapefile" contained in the biketrails_arc/ folder and load
# it into R using the sf package's read_sf() function.
bike_trails <- read_sf("biketrails_arc/") |> # If this doesn't run check the path to your file. 
  st_transform(crs = 4326)|>
  st_tran

# Plot this data!
ggplot() +
  geom_sf(data = bike_trails, color = "blue")

# The above map lacks context. Where are these bikepaths?
# Use ma_counties from problem 1 above obtain a map of Massachusetts counties.

# Plot this data!
ggplot() +
  geom_sf(data = ma_counties)

# Plot both maps by adding both layers!
ggplot() +
  geom_sf(data = bike_trails, color = "blue") + 
  geom_sf(data = ma_counties) 

# Where did our bike paths go? The order you plot in matters! Switch the two
# geom_sf()
ggplot() +
  geom_sf(data = ma_counties) +
  geom_sf(data = bike_trails, color = "blue")

# Better! But what are these counties? Let's mark each counties' centroid
ma_counties <- ma_counties |>
  # Add centroids to each region using purrr package.
  mutate(
    lng = purrr::map_dbl(geom, ~st_centroid(.x)[[1]]),
    lat = purrr::map_dbl(geom, ~st_centroid(.x)[[2]])
    )
ma_counties 

# Plot! Notice how we define data = in the geom_point() instead of the ggplot().
# This sets the data set only for the geom_point() and not the other geoms.
# Any data = defined in the global ggplot() trickles down by default to the other
# geoms.
ggplot() +
  geom_sf(data = ma_counties, size = 0.5) +
  geom_sf(data = bike_trails, color = "blue") +
  geom_point(data = ma_counties, aes(x = lng, y = lat)) 

# Can we do better? Yes! Use geom_text() instead with the label aesthetic!
ggplot() +
  geom_sf(data = ma_counties, size = 0.5) +
  geom_sf(data = bike_trails, color = "blue") +
  geom_text(data = ma_counties, aes(x = lng, y = lat, label = county_name)) 
```

## Exercise

Part 1: Create a map of the state of Vermont showing the top 9 ski resorts by Acres.

Part 2: Some of the resorts have leases on public land. The shape files are located in the Alpine_Ski_Area_Leases folder and were downloaded from https://geodata.vermont.gov/search?collection=Dataset. Add these to your map and color them blue.

> Hint: You get to get the coordinates using geocode(), then set `crs` , before using st_as_sf()

```{r}
#| eval: False

# Nothing to do here I scraped and cleaned up the resorts list for you, mostly because the server... 

# Write your code below
ski_resorts <- "https://bestmapsever.com/pages/vermont-ski-areas-resorts-list"

# This scraps the names, Acres, Vertical, Trails, and List from the website above. Use the slice_max() function to grab the top 10 resorts by acres.   

vt_ski_resorts <- ski_resorts |>
  read_html() |> 
  html_element("table") |>
  html_table() |>
  clean_names() |>
  slice_max(order_by = acres, n = 10)

# geocode the addresses and save them in the vt_ski_resorts_df. wait 1 second per address.
vt_ski_resorts <- vt_ski_resorts |>
  geocode(resort, method="osm")

write_csv("vt_ski_resorts.csv")
```

```{r}
# Complete part 1 here. 

vt_ski_resorts <- read_csv("vt_ski_resorts.csv") |> slice_max(order_by = acres, n = 9)

# transfrom your data to have the correct projection.  
vt_ski_resorts <- vt_ski_resorts |>
  st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE)

vt_counties <- maps::map("state",regions = "vermont",  plot = FALSE, fill = TRUE) |>
  # This sets the simple features 
  st_as_sf(coords = c("x", "y"), crs = 4326) |>
  # This adds a new column with the names of the counties. 
  mutate(county_name = str_replace(ID,pattern = "vermont,", replace = ""))

# Load in the lease shape file. 
leases <- read_sf("Alpine_Ski_Area_Leases/") |>
  st_transform(crs = 4326)

# plot the vermont ski areas. When plotting the ski resorts, set shape = 8  for a snowflake. Add the name with geom_text, set the size to 2 and nudge_x to move the names to the right.  

ggplot()+
  geom_sf(data = vt_counties, color = "green")+
  geom_sf(data = leases, color="blue")+
  geom_sf(data = vt_ski_resorts, shape = 8, )+
  geom_text(data = vt_ski_resorts, aes(x=long, y=lat, label= resort), size = 2, nudge_x = 0.75)+
  theme_minimal()+
  theme(
    axis.text = element_blank(),
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
  ) +
  labs(
    title = "Top 9 Aki Areas VT",
    subtitle = "Resort Land leases in blue",
    x="",
    y=""
  )


```

Part 3: Go back and get the top 10 ski resorts in Vermont, instead of 9 and redraw your map, what do you notice? Why do you think this happened? How do you think we can get Pico on the VT map correctly? Try it below to make a map of the top 10 resorts.

```{r}
vt_ski_resorts <- read_csv("vt_ski_resorts.csv") |> slice_max(order_by = acres, n = 10)

# plot the vermont ski areas. When plotting the ski resorts, set shape = 8  for a snowflake. Add the name with geom_text, set the size to 2 and nudge_x to move the names to the right.  

ggplot()+
  geom_sf(data = vt_counties, color = "green")+
  geom_sf(data = leases, color="blue")+
  geom_sf(data = vt_ski_resorts, shape = 8, )+
  geom_text(data = vt_ski_resorts, aes(x=long, y=lat, label= resort), size = 2, nudge_x = 0.75)+
  theme_minimal()+
  theme(
    axis.text = element_blank(),
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
  ) +
  labs(
    title = "Top 10 Ski Areas VT",
    subtitle = "Resort Land leases in blue",
    x="",
    y=""
  )


```

## Choropleth Maps

```{r}
library(tidyverse)
library(sf)
```

```{r}
mass_under_5 = read_sf("mass_under_5")
```

```{r}
# Pop of mass

ggplot()+
  geom_sf(data = mass_under_5, aes( fill = pop))+
  scale_fill_distiller(direction = 1) + # dark is more
  theme_void()+
  labs( 
    title = "Population of Massachusetts by County",
    caption = "America Community Survey, 2024")

```

```{r}
ggplot()+
  geom_sf(data = mass_under_5, aes( fill = tot_udr_5))+
  scale_fill_distiller(direction = 1) + # dark is more
  theme_void()+
  labs( 
    title = "Total Under 5 y.o. by County",
    caption = "America Community Survey, 2024")


```

```{r}
mass_under_5 = mass_under_5 |>
  mutate(prop = tot_udr_5/pop)

ggplot()+
  geom_sf(data = mass_under_5, aes( fill = prop))+
  scale_fill_distiller(direction = 1) + # dark is more
  theme_void()+
  labs( 
    title = "Prop Under 5 y.o. by County",
    caption = "America Community Survey, 2024")

  
```

```{r}
mass_retiree = read_sf("mass_retiree")

ggplot()+
  geom_sf(data = mass_retiree, aes( fill = pop))+
  scale_fill_distiller(direction = 1) + # dark is more
  theme_void()+
  labs( 
    title = "People Over 65 by County",
    caption = "America Community Survey, 2024")

mass_retiree = read_sf("mass_retiree")
```

```{r}
ggplot()+
  geom_sf(data = mass_retiree, aes( fill = rtr_prp))+
  scale_fill_distiller(direction = 1) + # dark is more
  theme_void()+
  labs( 
    title = "Prop Over 65 by County",
    caption = "America Community Survey, 2024")

mass_retiree = mass_retiree |> rename(pop = retiree )
```

```{r}
ggplot()+
  geom_sf(data = mass_retiree, aes( fill = rtr_prp*pop))+
  scale_fill_distiller(direction = 1) + # dark is more
  theme_void()+
  labs( 
    title = "People Over 65 by County",
    caption = "America Community Survey, 2024")

```
